---
title: "Obstetrics and Gynecology (ObGyn) Cervical Tissue Stiffness Study"
author: Molly Stout, Peinan Zhao, ... , Simeone Marino, Mingyi Tang, Yang Chen, Ivo D. Dinov
date: "`r format(Sys.time(), '%D %B %Y')`"
output:
  html_document: 
    theme: spacelab
    highlight: tango
    toc: yes
    number_sections: yes
    toc_depth: 4
    toc_float:
      collapsed: no
      smooth_scroll: yes
tags:
- DSPA
- SOCR
- OBGYN
- Predictive Analytics
subtitle: "Fully Quantitative Cervical Elastography System (FQ-CES) - Modeling, Validation & Analytics"
always_allow_html: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE, echo=TRUE, message=FALSE, warning=FALSE)
library(dplyr)
library(reshape2)
library(rmarkdown)
```


# Data Wrangling, Imputation and Rebalancing

ADD SOME TEXT HERE ON THE PROTOCOL USED

```{r eval=FALSE}
#setwd(file.path("C:/Users/11191/Desktop/2022Fall/MDP/SOCR_OBGYN_MollyStout_AncaTilea_2022/Pilot_data"))
data = read.csv("FullyQuantitativeCer_DATA_2022-11-02_0853.csv", stringsAsFactors=TRUE)
ym = read.csv("logYM_Updated.csv", stringsAsFactors=TRUE) %>%
  distinct(record_id, pvisit, .keep_all = TRUE)
#dim(data)
#dim(data_um)
# Hmisc::describe(data)
```

Calculation before building the new data frame

```{r eval=FALSE}
data_general = data[!is.na(data$a1_consented), ]
data_general = data_general[data_general$redcap_event_name == "general_arm_1" &
                              data_general$a1_consented==1, ]
data_general$delivery_ga = 
  data_general$delivery_ga_weeks * 7 + data_general$delivery_ga_days
data_general$preterm_birth[
  is.na(data_general$preterm_birth) | data_general$preterm_birth == 3] =
  as.numeric(
    !(data_general$delivery_ga_weeks[
      is.na(data_general$preterm_birth) | 
        data_general$preterm_birth == 3] < 37)) + 1
data_general$bmi = data_general$bmi_1st_visit
data_general$education_level = data_general$education
data_general$insurance_status = data_general$insurance
data_general$preterm_history = data_general$preterm > 0
data_general$age = as.numeric(difftime(data_general$edc, 
                                       data_general$dob) / 365)
data_general = data_general[c("record_id",
                              "delivery_ga",
                              "preterm_birth",
                              "race", 
                              "age", 
                              "bmi",
                              "education_level",
                              "employed",
                              "insurance_status",
                              "preterm_history",
                              "sptb",
                              "term",
                              "preterm",
                              "prior_vaginal_deliveries",
                              "number_prior_cs")]
data_visit = data[data$redcap_event_name != "general_arm_1",
                  c("record_id", 
                    "redcap_event_name", 
                    "cl_v1", 
                    "ga_wks_a1", 
                    "ga_days_a1")]
data_visit = data_visit[data_visit$record_id %in% data_general$record_id, ]
data_visit$ga_wks_a1 = as.numeric(as.character(data_visit$ga_wks_a1))
data_visit$ga_days_a1 = as.numeric(as.character(data_visit$ga_days_a1))
data_visit$ga = 7 * data_visit$ga_wks_a1 + data_visit$ga_days_a1
data_visit = left_join(data_visit, ym, 
                       by=c("record_id" = "record_id", "ga" = "pvisit"))

data_visit = dcast(
  data_visit[c("record_id", "redcap_event_name", "cl_v1")], 
  record_id ~ redcap_event_name) %>%
  inner_join(dcast(data_visit[c("record_id", "redcap_event_name", "ga")], 
                   record_id ~ redcap_event_name),
             by="record_id") %>%
  inner_join(dcast(data_visit[c("record_id", "redcap_event_name", "PredYM")], 
                   record_id ~ redcap_event_name),
             by="record_id")
colnames(data_visit)[2:10] = c("cl_1", "cl_2", "cl_3",
                               "ga_1", "ga_2", "ga_3",
                               "PredYM_1", "PredYM_2", "PredYM_3")
data_visit = left_join(data_general, data_visit, by="record_id")
```

```{r eval=FALSE}
library(missForest)

data = data_visit

data = data[!is.na(data$preterm_birth) & data$preterm_birth != 3, ]
delivery_ga = data$delivery_ga
data = select(data, -delivery_ga)
data$preterm_birth = as.factor(-data$preterm_birth + 2)
preterm_birth = data$preterm_birth
data = select(data, -preterm_birth)

data$race = as.factor(data$race)

data$employed = as.factor(data$employed)

data$insurance_status = as.factor(data$insurance_status)

data$preterm_history = as.factor(as.numeric(data$preterm_history))

data$sptb = as.factor(-data$sptb + 2)
sptb = data$sptb
data = select(data, -sptb)

data$prior_vaginal_deliveries[data$term + data$preterm == 0] = 0
data$number_prior_cs[data$term + data$preterm == 0] = 0
data$prior_vaginal_deliveries[is.na(data$prior_vaginal_deliveries)] = 
  (data$preterm + data$term
   - data$number_prior_cs)[is.na(data$prior_vaginal_deliveries)]
data$number_prior_cs[is.na(data$number_prior_cs)] = 
  (data$preterm + data$term 
   - data$prior_vaginal_deliveries)[is.na(data$number_prior_cs)]

data[unlist(lapply(data, is.numeric))][
  data[unlist(lapply(data, is.numeric))] < 0] = NA

rownames(data) = data$record_id
data = select(data, -record_id)

data = missForest(data, verbose=T)$ximp

data$cl_fd_1 = (data$cl_1 - data$cl_2) / data$cl_1
data$cl_fd_2 = (data$cl_2 - data$cl_3) / data$cl_2
data$PredYM_fd_1 = (data$PredYM_1 - data$PredYM_2) / data$PredYM_1
data$PredYM_fd_2 = (data$PredYM_2 - data$PredYM_3) / data$PredYM_2
data$change_1 = data$ga_2 - data$ga_1
data$change_2 = data$ga_3 - data$ga_2

data_um = data[grepl("20-", rownames(data)), ]
delivery_ga_um = delivery_ga[grepl("20-", rownames(data))]
preterm_birth_um = preterm_birth[grepl("20-", rownames(data))]
sptb_um = sptb[grepl("20-", rownames(data))]
```

```{r eval=FALSE}
rm(data_general)
#rm(data_visit)
rm(ym)
save.image("predict_preterm_data.RData")
```

These are the unbalanced/original dataset dimensions

```{r}
load("predict_preterm_data.RData")
#dim(data_um)
table(preterm_birth_um)

```

# Survival Analysis

## Kaplan-Meier Method

```{r}
library(survival)

m1 = survfit(Surv(delivery_ga_um, rep(1, length(delivery_ga_um))) ~ 1)
summary(m1)
plot(m1)
plot(m1$time, -log(m1$surv), type="l")
plot(log(m1$time), log(-log(m1$surv)), type="l")
```

Log Survival Plot: Concave up, the probability of delivery is increasing over time.

Since both plots are not straight lines, time to deliver neither follows an Exponential distribution nor a Weibull distribution.

### Effect of whether theres is a third visit of a patient

```{r}
model_ga_3 = survfit(
  Surv(delivery_ga, rep(1, length(delivery_ga))) ~ is.na(ga_3),
  data=data_visit)
plot(model_ga_3, col=c(3, 4), lty=1:2)
survdiff(
  Surv(delivery_ga, rep(1, length(delivery_ga))) ~ is.na(ga_3),
  data=data_visit)
survdiff(
  Surv(delivery_ga, rep(1, length(delivery_ga))) ~ is.na(ga_3),
  data=data_visit, rho=1)
legend(10, 0.8, c("FALSE", "TRUE"), col=c(3, 4), lty=1:2)
```


## Parametric Regression Model

```{r}
library(ggplot2)
ggplot(NULL, aes(x=delivery_ga_um)) + 
  geom_histogram(aes(y=..density..), color="black", fill="white", bins=100) +
  geom_density(alpha=.2, fill="#FF6666") +
  scale_x_log10()
```

### Log-Normal disribution

```{r}
data_surv = cbind(delivery_ga_um, data_um)

linear_fit = survreg(
  Surv(delivery_ga_um, rep(1, length(delivery_ga_um))) ~ ., 
  data = data_surv,
  dist="lognormal")
summary(linear_fit)

library(car)
qqPlot(residuals(linear_fit))
```

Interpretation Example:

The signs of the coefficients tell us the direction of the relationship.

The positive coefficient for `preterm_history1` indicates that those who have preterm history have longer time to deliver than those who did not. If we calculate $e^{\hat{\beta}}$, we get the estimated ratio of the expected gestational age of delivery for the two groups. In the case, the estimation is $e^{0.024144} = 1.027841$. Therefore, controlling for other predictors in the model, the expected time to deliver for those who have preterm history is $2.7841\%$ higher than those who do not have preterm history.

The negative coefficient for `race6` indicates that those who are Hispanic or Latino have shorter time to deliver.

### Extreme-Value distribution

```{r}
extreme_value_dist = survreg(
  Surv(delivery_ga_um, rep(1, length(delivery_ga_um))) ~ 
    age + bmi + term + preterm + 
    cl_1 + cl_2 + cl_3 + ga_1 + ga_2 + ga_3 + PredYM_1 + PredYM_2 + PredYM_3,
  data = data_surv,
  dist="extreme")
summary(extreme_value_dist)

qqPlot(residuals(extreme_value_dist))
```

Interpretation example:

For a quantitative variable such as `ga_3`, we exponentiate the coefficient estimator to get the increase in the expected time to deliver for a one-unit increase in the variable.

`term` has a negative coefficient, so the estimation is $e^{-2.2997} = 0.1002889$. According to the model then, a one-unit increase in `term` is associated with a $1 - 0.1002889 = 89.97111\%$ percent decrease in expected time to deliver, holding other predictors constant.

## Proportional Hazards Model

```{r}
proportional_hazards_model = coxph(
  Surv(delivery_ga_um, rep(1, length(delivery_ga_um))) ~ 
    age + bmi + term + preterm + 
    cl_1 + cl_2 + cl_3 + ga_1 + ga_2 + ga_3 + PredYM_1 + PredYM_2 + PredYM_3,
  data = data_surv
)
summary(proportional_hazards_model)

qqPlot(residuals(proportional_hazards_model))
```

Interpretation example:

For a one unit increase in the variable `cl_3`, the hazard ratio is $e^{-0.031914} = 0.9685899$. This means that the hazard of delivery goes down by an estimated $3.14101\%$.

<!--

# CBDA with original data

```{r cache=FALSE, eval=FALSE}
library(SuperLearner)

#randomForest_learners = 
#  create.Learner(
#    "SL.randomForest",
#    tune=list(mtry=floor(sqrt(ncol(x_train)) * c(0.25, 0.5, 1, 2, 4)),
#              ntree=nrow(x_train) * c(0.125, 0.25, 0.5, 1, 2, 4)))

#ranger_learners = 
#  create.Learner(
#    "SL.ranger",
#    tune=list(num.trees=nrow(x_train) * c(0.125, 0.25, 0.5, 1, 2, 4),
#              mtry = floor(sqrt(ncol(x_train)) * c(0.25, 0.5, 1, 2, 4))))

#xgboost_learners = 
#  create.Learner(
#    "SL.xgboost",
#    tune=list(ntrees=nrow(x_train) * c(0.125, 0.25, 0.5, 1, 2, 4),
#              max_depth = c(4, 8, 16),
#              shrinkage = c(0.025, 0.05, 0.1, 0.2)))


model = function(X, Y, newX, newY, features=NULL) {
  if (is.null(features)) {
    features = colnames(X)
  }
  sl = SuperLearner(
    Y=Y,
    X=X,
    newX=newX,
    family=binomial(),
    cvControl=list(stratifyCV=TRUE),
    SL.library=c(#"SL.cforest",
                 "SL.gbm", 
                 "SL.knn",
                 "SL.ksvm", 
                 #randomForest_learners$names,
                 #ranger_learners$names,
                 #xgboost_learners$names,
                 "SL.randomForest",
                 "SL.ranger",
                 "SL.xgboost"),
    verbose=TRUE)
  pred_rocr = ROCR::prediction(sl$SL.predict, newY)
  auc = ROCR::performance(
    pred_rocr, measure="auc", x.measure="cutoff")@y.values[[1]]
  y_pred = factor(as.numeric(sl$SL.predict > 0.5), levels=c(0, 1))
  stats = caret::confusionMatrix(y_pred, as.factor(newY))
  return(list(sl=sl, auc=auc, stats=stats, y_pred=y_pred))
}
```

## Predictive model on the original unbalanced data including all the features (internal validation only)

```{r, eval=FALSE}
set.seed(1)
model_unbalance = model(data_um, 
                        as.numeric(as.character(preterm_birth_um)),
                        data_um,
                        as.numeric(as.character(preterm_birth_um)))
model_unbalance$sl
model_unbalance$auc
model_unbalance$stats
```


```{r, eval=FALSE}
library(DT)

validation_result = data.frame(matrix(ncol=0, nrow=length(preterm_birth_um)))
rownames(validation_result) = rownames(data_um)
validation_result$ground_truth = as.numeric(as.character(preterm_birth_um))

for (i in 1:20) {
  set.seed(i)
  model_ex_valid = model(data_um, 
                         as.numeric(as.character(preterm_birth_um)),
                         data_um,
                         as.numeric(as.character(preterm_birth_um)))
  validation_result = cbind(validation_result, 
                            as.numeric(as.character(model_ex_valid$y_pred)))
}

miss_classify = data.frame(
  apply(abs(validation_result[2:dim(validation_result)[2]] -
              validation_result$ground_truth), 1, sum))
miss_classify = cbind(miss_classify,
                      as.numeric(as.character(preterm_birth_um)))
colnames(miss_classify) = c("miss_classify_count", "ground_truth")

datatable(miss_classify, filter="top",
          options=list(pageLength=10, autoWidth=TRUE))
```



# CBDA with rebalanced data

These are the rebalanced dataset dimensions


```{r, eval=FALSE}
library(DMwR)

data_um_balanced = SMOTE(preterm_birth_um ~ ., 
                         cbind(preterm_birth_um, data_um), 
                         perc.over=400, perc.under=125)
preterm_birth_um_balanced = data_um_balanced$preterm_birth_um
data_um_balanced = select(data_um_balanced, -preterm_birth_um)
preterm_birth_um_balanced = as.numeric(as.character(preterm_birth_um_balanced))

dim(data_um_balanced)
table(preterm_birth_um_balanced)
```

## Predictive model on the rebalanced data including all the features (internal validation only) - POSITIVE CONTROL

```{r, eval=FALSE}
x_train = data_um_balanced
x_valid = x_train
y_train = preterm_birth_um_balanced
y_valid = y_train

model_unbalance = model()
model_unbalance$sl
model_unbalance$auc
model_unbalance$stats
```

## Predictive model including all the features

We rebalanced the data so we can test a predictive model with external validation data (training/validation, either 80%/20% or 90%/10%).

```{r, eval=FALSE}
held_up = 0.8
print(held_up)

train_obs = sample(nrow(data_um_balanced), 
                   round(nrow(data_um_balanced) * held_up))
x_train = data_um_balanced[train_obs, ]
x_valid = data_um_balanced[-train_obs, ]
y_train = preterm_birth_um_balanced[train_obs]
y_valid = preterm_birth_um_balanced[-train_obs]
```

These are all the features included in the model

*race, age, bmi, education_level, employed, insurance_status, preterm_history, term, preterm, prior_vaginal_deliveries, number_prior_cs, cl_1, cl_2, cl_3, ga_1, ga_2, ga_3, logYM_1, logYM_2, logYM_3, cl_fd_1, cl_fd_2, logYM_fd_1, logYM_fd_2, range_1, range_2*


```{r, eval=FALSE}
model_all = model()
model_all$sl
model_all$auc
model_all$stats
```

## Predictive model including the features returned by Knockoff 

```{r, eval=FALSE}
library(knockoff)

filtered_features = names(
  knockoff.filter(
    data_um_balanced[unlist(lapply(data_um_balanced, is.numeric))], 
    preterm_birth_um_balanced, fdr=0.5)$selected)
filtered_features
```

```{r, eval=FALSE}
model_ko = model(filtered_features)
model_ko$sl
model_ko$auc
model_ko$stats
```

## Predictive model including a subset of clinical features 

These are the features included: 

*race, age, bmi, education_level, employed, insurance_status, and preterm_history*.

```{r, eval=FALSE}
model_cc = model(c("race", 
                   "age", 
                   "bmi", 
                   "education_level", 
                   "employed", 
                   "insurance_status", 
                   "preterm_history"))
model_cc$sl
model_cc$auc
model_cc$stats
```

```{r, eval=FALSE}
model_ym1 = model(c("logYM_1"))
model_ym1$sl
model_ym1$auc
model_ym1$stats
```

```{r, eval=FALSE}
model_ym2 = model(c("logYM_2"))
model_ym2$sl
model_ym2$auc
model_ym2$stats
```

```{r, eval=FALSE}
model_ymchange = model(c("logYM_fd_1"))
model_ymchange$sl
model_ymchange$auc
model_ymchange$stats
```

```{r, eval=FALSE}
model_cl1 = model(c("cl_1"))
model_cl1$sl
model_cl1$auc
model_cl1$stats
```

```{r, eval=FALSE}
model_cl2 = model(c("cl_2"))
model_cl2$sl
model_cl2$auc
model_cl2$stats
```

```{r, eval=FALSE}
model_cc_ym1 = model(c("race", 
                       "age", 
                       "bmi", 
                       "education_level", 
                       "employed", 
                       "insurance_status", 
                       "preterm_history",
                       "logYM_1"))
model_cc_ym1$sl
model_cc_ym1$auc
model_cc_ym1$stats
```

```{r, eval=FALSE}
model_cc_ym2 = model(c("race", 
                       "age", 
                       "bmi", 
                       "education_level", 
                       "employed", 
                       "insurance_status", 
                       "preterm_history",
                       "logYM_2"))
model_cc_ym2$sl
model_cc_ym2$auc
model_cc_ym2$stats
```

```{r, eval=FALSE}
model_cc_ymchange = model(c("race", 
                            "age", 
                            "bmi", 
                            "education_level", 
                            "employed", 
                            "insurance_status", 
                            "preterm_history",
                            "logYM_fd_1"))
model_cc_ymchange$sl
model_cc_ymchange$auc
model_cc_ymchange$stats
```

```{r, eval=FALSE}
model_cc_cl1 = model(c("race", 
                       "age", 
                       "bmi",
                       "education_level", 
                       "employed", 
                       "insurance_status", 
                       "preterm_history",
                       "cl_1"))
model_cc_cl1$sl
model_cc_cl1$auc
model_cc_cl1$stats
```

```{r, eval=FALSE}
model_cc_cl2 = model(c("race", 
                       "age", 
                       "bmi",
                       "education_level", 
                       "employed", 
                       "insurance_status", 
                       "preterm_history",
                       "cl_2"))
model_cc_cl2$sl
model_cc_cl2$auc
model_cc_cl2$stats
```

```{r, eval=FALSE}
model_final = model(c("race", 
                      "age", 
                      "bmi",
                      "education_level", 
                      "employed", 
                      "insurance_status", 
                      "preterm_history",
                      "cl_1",
                      "cl_2",
                      "logYM_1",
                      "logYM_2",
                      "logYM_fd_1"))
model_final$sl
model_final$auc
model_final$stats
```

-->
